---
title: "CFP parcels with major canopy loss"
subtitle: "Houghton & Keweenaw Counties, Michigan"
author: "Randy Swaty"
date: today
format:
  html:
    embed-resources: true
    page-layout: full
    fontsize: 0.9em
    theme: flatly
    toc: false
    fig-width: 10
    fig-height: 5
    fig-dpi: 140
  pdf:
    documentclass: article
    geometry:
      - margin=0.7in
    fontsize: 10pt
    toc: false
    fig-width: 6.2
    fig-height: 2.8
    fig-pos: "H"
    keep-tex: false
    include-in-header:
      - text: |
          \usepackage{float}
          \usepackage[section]{placeins}
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(dplyr)
library(ggplot2)
library(sf)
library(scales)
library(knitr)

options(scipen = 999)

# Whole numbers with commas; never scientific notation.
fmt_n <- function(x) {
  format(round(as.numeric(x)), big.mark = ",", scientific = FALSE, trim = TRUE)
}

root <- if (dir.exists("data/processed")) "." else ".."
rank <- read.csv(
  file.path(root, "data/processed/cfp_parcel_tcc_change_by_owner_type.csv"),
  stringsAsFactors = FALSE
)

p20 <- st_read(file.path(root, "data/cfp_data/cfp_hk_2020.shp"), quiet = TRUE) |>
  st_drop_geometry() |>
  transmute(
    parid = as.character(parid),
    name_2020 = trimws(as.character(search_nam))
  )
p26 <- st_read(file.path(root, "data/cfp_data/cfp_hk_2026.shp"), quiet = TRUE) |>
  st_drop_geometry() |>
  transmute(
    parid = as.character(parid),
    name_2026 = trimws(as.character(FullLegalN))
  ) |>
  distinct(parid, .keep_all = TRUE)

norm <- function(s) {
  s <- tolower(as.character(s))
  s <- gsub("[^a-z0-9 ]", " ", s)
  s <- gsub("\\b(llc|inc|ltd|co|company|corp|corporation)\\b", "", s)
  gsub("\\s+", " ", trimws(s))
}

loss <- rank |>
  filter(delta_pp < -20) |>
  mutate(parid = as.character(parid)) |>
  left_join(p20, by = "parid") |>
  left_join(p26, by = "parid") |>
  mutate(
    type_changed = owner_type_2020 != owner_type_2026 &
      owner_type_2026 != "non-CFP in 2026",
    name_same_loose = norm(name_2020) == norm(name_2026) & nzchar(norm(name_2020)),
    name_from = case_when(
      grepl("TRG THRESHOLD", name_2020, ignore.case = TRUE) ~ "TRG Threshold*",
      grepl("KEWEENAW LAND", name_2020, ignore.case = TRUE) ~ "Keweenaw Land Assoc*",
      grepl("MWF", name_2020, ignore.case = TRUE) ~ "MWF*",
      TRUE ~ "Other"
    ),
    name_to = case_when(
      grepl("Verdant", name_2026, ignore.case = TRUE) ~ "Verdant Timber*",
      grepl("Sage Timber", name_2026, ignore.case = TRUE) ~ "Sage Timber*",
      grepl("MWF", name_2026, ignore.case = TRUE) ~ "MWF*",
      TRUE ~ "Other"
    ),
    size_bin = cut(
      acres_2020,
      breaks = c(0, 40, 80, 160, 320, Inf),
      labels = c("20–40", "40–80", "80–160", "160–320", ">320"),
      right = TRUE
    )
  )

n_all <- nrow(rank)
n_loss <- nrow(loss)
ac_loss <- sum(loss$acres_2020)
n_fi_stable <- sum(loss$owner_type_2020 == "Forest Industry" &
  loss$owner_type_2026 == "Forest Industry")
n_type_chg <- sum(loss$type_changed)
n_name_chg <- sum(!loss$name_same_loose)
ac_name_chg <- sum(loss$acres_2020[!loss$name_same_loose])
pct_loss <- round(100 * n_loss / n_all)
med_abs <- median(loss$abs_delta_pp)
max_abs <- max(loss$abs_delta_pp)
ac_trg_verdant <- sum(
  loss$acres_2020[loss$name_from == "TRG Threshold*" & loss$name_to == "Verdant Timber*"]
)
ac_kla_sage <- sum(
  loss$acres_2020[loss$name_from == "Keweenaw Land Assoc*" & loss$name_to == "Sage Timber*"]
)
med_parcel_ac <- median(loss$acres_2020)
mean_parcel_ac <- mean(loss$acres_2020)
# Heartlands / TNC among major-loss set (usually none; full TNC exit is in sellers/keepers doc)
n_tnc_loss <- sum(grepl("Nature Conservancy", loss$name_2026, ignore.case = TRUE))
ac_tnc_loss <- sum(
  loss$acres_2020[grepl("Nature Conservancy", loss$name_2026, ignore.case = TRUE)]
)

theme_one <- theme_minimal(base_size = 10) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 11),
    plot.subtitle = element_text(color = "grey35", size = 8.5),
    axis.title = element_text(size = 9),
    legend.position = "bottom",
    legend.title = element_text(size = 8),
    legend.text = element_text(size = 8),
    plot.margin = margin(4, 4, 4, 4)
  )

label_whole <- label_comma(accuracy = 1, big.mark = ",")
```

### Summary

Using Commercial Forest Parcel (CFP) data for 2020 and 2026 from the Michigan Department of Natural Resources alongside the USFS FIA Total Canopy Cover (TCC) dataset across Houghton and Keweenaw counties, I screened for parcels with mean canopy cover change steeper than −20 percentage points between 2020 and 2025.^[The −20 pp cut is a **screening threshold** for “major” parcel-mean loss in this memo—not an industry biodiversity standard or a claim that every such change is harvest rather than other disturbance. Edge spill from neighboring heavy-loss parcels can push a parcel’s mean below −20 pp even when interior cover is steadier; parcels adjacent to very high-loss neighbors with smaller mean changes are outside this set.] All of the information below is for that major-loss subset.

Of `{r} fmt_n(n_all)` CFP parcels ≥20 acres, **`{r} fmt_n(n_loss)` parcels (~`{r} fmt_n(ac_loss)` acres)** lost more than **20 percent** of mean tree canopy from 2020 to 2025. Almost all kept the state **Forest Industry** owner-type label^[Michigan DNR assigns CFP `owner_type` / `OwnerTypeT` categories (including Forest Industry); it is a program classification, not our institutional-owner roster. See `docs/cfp_institutional_sellers_keepers.qmd` for TRG / Lyme / MWF contrasts.] — but most **changed legal name**, with the most common change being TRG Threshold → Verdant Timber.

### Key numbers

| | |
|:---|---:|
| Parcels with ΔTCC < −20 percent | **`{r} fmt_n(n_loss)`** (`{r} fmt_n(pct_loss)`% of ranked parcels) |
| Acres (2020 attribute) | **`{r} fmt_n(ac_loss)`** |
| Median / mean parcel size (2020 acres) | **`{r} fmt_n(med_parcel_ac)`** / `{r} fmt_n(mean_parcel_ac)` |
| Still Forest Industry → Forest Industry | **`{r} fmt_n(n_fi_stable)`** parcels |
| Owner **type** changed | **`{r} fmt_n(n_type_chg)`** |
| Owner **name** changed (loose match) | **`{r} fmt_n(n_name_chg)`** (~`{r} fmt_n(ac_name_chg)` ac) |
| Left CFP by 2026 | **0** of these `{r} fmt_n(n_loss)` |
| Median / max |Δ| among the `{r} fmt_n(n_loss)` | `{r} fmt_n(med_abs)` / `{r} fmt_n(max_abs)` percent |


### General Parcel Size Distribution

<br>

```{r}
#| label: fig-size
#| fig-cap: "Size of the 72 major-loss parcels."
size_df <- loss |>
  group_by(size_bin) |>
  summarise(n = n(), acres = sum(acres_2020), .groups = "drop")

ggplot(size_df, aes(size_bin, acres)) +
  geom_col(fill = "#2f5d3a", width = 0.72) +
  geom_text(aes(label = paste0(fmt_n(n), " parcels")), vjust = -0.35, size = 2.8) +
  scale_y_continuous(labels = label_whole, expand = expansion(mult = c(0, 0.18))) +
  labs(
    title = "Acreage by parcel size",
    subtitle = "Ten parcels >320 ac hold nearly half the major-loss acreage",
    x = "Parcel acres (2020)",
    y = "Total acres"
  ) +
  theme_one
```


### Parcel type and owner transitions for parcels with greatest canopy cover change


```{r}
#| label: fig-type
#| fig-cap: "Ownership type continuity among major-loss parcels."


type_df <- loss |>
  count(owner_type_2020, owner_type_2026, name = "n") |>
  mutate(flow = reorder(paste(owner_type_2020, "→", owner_type_2026), n))

ggplot(type_df, aes(n, flow)) +
  geom_col(fill = "#4a7c59", width = 0.7) +
  geom_text(aes(label = fmt_n(n)), hjust = -0.15, size = 2.8) +
  scale_x_continuous(labels = label_whole, expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Owner type 2020 → 2026",
    x = "Parcels",
    y = NULL
  ) +
  theme_one


```


<br>


```{r}
#| label: fig-delta-owner
#| fig-cap: "Ownership name continuity among major-loss parcels."


name_df <- loss |>
  group_by(name_from, name_to) |>
  summarise(n = n(), acres = sum(acres_2020), .groups = "drop") |>
  mutate(flow = reorder(paste(name_from, "→", name_to), acres))

ggplot(name_df, aes(acres, flow)) +
  geom_col(fill = "#8b5a2b", width = 0.7) +
  geom_text(
    aes(label = paste0(fmt_n(n), " · ", fmt_n(acres))),
    hjust = -0.05,
    size = 2.6
  ) +
  scale_x_continuous(labels = label_whole, expand = expansion(mult = c(0, 0.28))) +
  labs(
    title = "Owner name 2020 → 2026",
    x = "Acres",
    y = NULL
  ) +
  theme_one
```

### Canopy cover change

```{r}
#| label: fig-delta-type
#| fig-cap: "Distribution of canopy change (percent) for major-loss parcels."
#| fig-pos: "H"
ggplot(loss, aes(delta_pp)) +
  geom_histogram(binwidth = 5, fill = "#1f4e3a", color = "white", linewidth = 0.3) +
  geom_vline(xintercept = -20, linetype = 2, color = "grey40") +
  scale_x_continuous(breaks = seq(-75, -20, by = 10), labels = label_whole) +
  scale_y_continuous(labels = label_whole) +
  labs(
    title = "How large were the losses?",
    subtitle = "Mean parcel TCC (percent); dashed line = −20 percent threshold",
    x = "Change in canopy (2025 − 2020), percent",
    y = "Parcels"
  ) +
  theme_one
```

```{=latex}
\FloatBarrier
```

### Takeaway

The ~`{r} fmt_n(ac_loss)` acres of **major canopy loss** are concentrated on CFP land that kept the **Forest Industry** type label. Continuity of owner *type* masks substantial **legal-name turnover**, especially **TRG Threshold Timber → Verdant Timber** (~55 parcels / ~`{r} fmt_n(ac_trg_verdant)` ac) and **Keweenaw Land Association → Sage Timber** (8 parcels / ~`{r} fmt_n(ac_kla_sage)` ac). None of these `{r} fmt_n(n_loss)` parcels left the CFP between 2020 and 2026.

**Institutional owners vs Forest Industry.** TRG (and Verdant), Lyme, and MWF are the large **institutional** names in this geography; they usually appear under the state’s Forest Industry type. Stars in the name chart (`TRG Threshold*`, `MWF*`, etc.) mark **collapsed name groups** (several legal entities rolled up), not a quality flag. **MWF** is the CFP legal-name token for MWF Ned Lake / MWF V Vulcan entities (Molpus-family stand-in in the sellers/keepers analysis; no literal “Molpus” or “Manulife” strings in HK CFP 2020).

**Heartlands / TNC.** TRG → Nature Conservancy transfers are largely **outside** this major-loss subset (`{r} fmt_n(n_tnc_loss)` of the `{r} fmt_n(n_loss)` major-loss parcels; ~`{r} fmt_n(ac_tnc_loss)` ac). By contrast, 18 of the TRG → Verdant major-loss parcels lie inside the **American Songbird** footprint, which Verdant sold in 2026 (after the CFP 2026 snapshot, so it still reads as Verdant here). See `docs/cfp_institutional_sellers_keepers.qmd` for the seller vs keeper comparison.

```{=latex}
\FloatBarrier
```

### Methods (short)

- **Unit:** 2020 CFP parcels (`parid`), ranked for all parcels with attribute acres ≥ 20.
- **Canopy:** zonal mean USFS/NLCD TCC on the 30 m grid, **2020 vs 2025** (latest annual TCC; ownership snapshot is 2026).
- **Major loss:** mean ΔTCC < −20 percent (`{r} fmt_n(n_loss)` of `{r} fmt_n(n_all)` ranked parcels); see footnote in Summary.
- **Ownership:** 2020 `owner_type` / `search_nam`; 2026 `OwnerTypeT` / `FullLegalN`, joined on `parid`.
- **Name match:** loose string compare (case, punctuation, LLC/Inc stripped). Source table: `data/processed/cfp_parcel_tcc_change_by_owner_type.csv` (`scripts/09_cfp_parcel_tcc_change.R`).
- **Related:** institutional sellers vs keepers disturbance rates — `scripts/10_institutional_sellers_vs_keepers.R`.
