---
title: "CFP parcels with major canopy loss"
subtitle: "Houghton & Keweenaw Counties, Michigan"
author: "Randy Swaty"
date: today
format:
  html:
    embed-resources: true
    page-layout: full
    fontsize: 0.9em
    theme: flatly
    toc: false
    fig-width: 10
    fig-height: 5
    fig-dpi: 140
  pdf:
    documentclass: article
    geometry:
      - margin=0.7in
    fontsize: 10pt
    toc: false
    fig-width: 6.2
    fig-height: 2.8
    fig-pos: "H"
    keep-tex: false
    include-in-header:
      - text: |
          \usepackage{float}
          \usepackage[section]{placeins}
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(dplyr)
library(ggplot2)
library(sf)
library(scales)
library(knitr)

options(scipen = 999)

# Whole numbers with commas; never scientific notation.
fmt_n <- function(x) {
  format(round(as.numeric(x)), big.mark = ",", scientific = FALSE, trim = TRUE)
}

root <- if (dir.exists("data/processed")) "." else ".."
rank <- read.csv(
  file.path(root, "data/processed/cfp_parcel_tcc_change_by_owner_type.csv"),
  stringsAsFactors = FALSE
)

p20 <- st_read(file.path(root, "data/cfp_data/cfp_hk_2020.shp"), quiet = TRUE) |>
  st_drop_geometry() |>
  transmute(
    parid = as.character(parid),
    name_2020 = trimws(as.character(search_nam))
  )
p26 <- st_read(file.path(root, "data/cfp_data/cfp_hk_2026.shp"), quiet = TRUE) |>
  st_drop_geometry() |>
  transmute(
    parid = as.character(parid),
    name_2026 = trimws(as.character(FullLegalN))
  ) |>
  distinct(parid, .keep_all = TRUE)

norm <- function(s) {
  s <- tolower(as.character(s))
  s <- gsub("[^a-z0-9 ]", " ", s)
  s <- gsub("\\b(llc|inc|ltd|co|company|corp|corporation)\\b", "", s)
  gsub("\\s+", " ", trimws(s))
}

loss <- rank |>
  filter(delta_pp < -20) |>
  mutate(parid = as.character(parid)) |>
  left_join(p20, by = "parid") |>
  left_join(p26, by = "parid") |>
  mutate(
    type_changed = owner_type_2020 != owner_type_2026 &
      owner_type_2026 != "non-CFP in 2026",
    name_same_loose = norm(name_2020) == norm(name_2026) & nzchar(norm(name_2020)),
    name_from = case_when(
      grepl("TRG THRESHOLD", name_2020, ignore.case = TRUE) ~ "TRG Threshold*",
      grepl("KEWEENAW LAND", name_2020, ignore.case = TRUE) ~ "Keweenaw Land Assoc*",
      grepl("MWF", name_2020, ignore.case = TRUE) ~ "MWF*",
      TRUE ~ "Other"
    ),
    name_to = case_when(
      grepl("Verdant", name_2026, ignore.case = TRUE) ~ "Verdant Timber*",
      grepl("Sage Timber", name_2026, ignore.case = TRUE) ~ "Sage Timber*",
      grepl("MWF", name_2026, ignore.case = TRUE) ~ "MWF*",
      TRUE ~ "Other"
    ),
    size_bin = cut(
      acres_2020,
      breaks = c(0, 40, 80, 160, 320, Inf),
      labels = c("20–40", "40–80", "80–160", "160–320", ">320"),
      right = TRUE
    )
  )

n_all <- nrow(rank)
n_loss <- nrow(loss)
ac_loss <- sum(loss$acres_2020)
n_fi_stable <- sum(loss$owner_type_2020 == "Forest Industry" &
  loss$owner_type_2026 == "Forest Industry")
n_type_chg <- sum(loss$type_changed)
n_name_chg <- sum(!loss$name_same_loose)
ac_name_chg <- sum(loss$acres_2020[!loss$name_same_loose])
pct_loss <- round(100 * n_loss / n_all)
med_abs <- median(loss$abs_delta_pp)
max_abs <- max(loss$abs_delta_pp)
ac_trg_verdant <- sum(
  loss$acres_2020[loss$name_from == "TRG Threshold*" & loss$name_to == "Verdant Timber*"]
)
ac_kla_sage <- sum(
  loss$acres_2020[loss$name_from == "Keweenaw Land Assoc*" & loss$name_to == "Sage Timber*"]
)

theme_one <- theme_minimal(base_size = 10) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 11),
    plot.subtitle = element_text(color = "grey35", size = 8.5),
    axis.title = element_text(size = 9),
    legend.position = "bottom",
    legend.title = element_text(size = 8),
    legend.text = element_text(size = 8),
    plot.margin = margin(4, 4, 4, 4)
  )

label_whole <- label_comma(accuracy = 1, big.mark = ",")
```

### Summary

Using Commercial Forest Parcel (CFP) data for 2020 and 2026 from the Michigan Department of Natural Resources alongside the USFS FIA Total Canopy Cover (TCC) dataset, I looked for parcels that had a canopy cover change > 20% between 2020 and 2026.  All of the information below is for parcels that had the largest change in canopy cover.

Of `{r} fmt_n(n_all)` CFP parcels ≥20 acres, **`{r} fmt_n(n_loss)` parcels (~`{r} fmt_n(ac_loss)` acres)** lost more than **20 percent** of mean tree canopy from 2020 to 2025. Almost all stayed **Forest Industry** by owner type — but most **changed legal name**, led by TRG Threshold → Verdant Timber.

### Key numbers

| | |
|:---|---:|
| Parcels with ΔTCC < −20 percent | **`{r} fmt_n(n_loss)`** (`{r} fmt_n(pct_loss)`% of ranked parcels) |
| Acres (2020 attribute) | **`{r} fmt_n(ac_loss)`** |
| Still Forest Industry → Forest Industry | **`{r} fmt_n(n_fi_stable)`** parcels |
| Owner **type** changed | **`{r} fmt_n(n_type_chg)`** |
| Owner **name** changed (loose match) | **`{r} fmt_n(n_name_chg)`** (~`{r} fmt_n(ac_name_chg)` ac) |
| Left CFP by 2026 | **0** of these `{r} fmt_n(n_loss)` |
| Median / max |Δ| among the `{r} fmt_n(n_loss)` | `{r} fmt_n(med_abs)` / `{r} fmt_n(max_abs)` percent |


### General Parcel Size Distribution

<br>

```{r}
#| label: fig-size
#| fig-cap: "Size of the 72 major-loss parcels."
size_df <- loss |>
  group_by(size_bin) |>
  summarise(n = n(), acres = sum(acres_2020), .groups = "drop")

ggplot(size_df, aes(size_bin, acres)) +
  geom_col(fill = "#2f5d3a", width = 0.72) +
  geom_text(aes(label = paste0(fmt_n(n), " parcels")), vjust = -0.35, size = 2.8) +
  scale_y_continuous(labels = label_whole, expand = expansion(mult = c(0, 0.18))) +
  labs(
    title = "Acreage by parcel size",
    subtitle = "Ten parcels >320 ac hold nearly half the major-loss acreage",
    x = "Parcel acres (2020)",
    y = "Total acres"
  ) +
  theme_one
```


### Parcel type and owner transitions for parcels with greatest canopy cover change


```{r}
#| label: fig-type
#| fig-cap: "Ownership type continuity among major-loss parcels."


type_df <- loss |>
  count(owner_type_2020, owner_type_2026, name = "n") |>
  mutate(flow = reorder(paste(owner_type_2020, "→", owner_type_2026), n))

ggplot(type_df, aes(n, flow)) +
  geom_col(fill = "#4a7c59", width = 0.7) +
  geom_text(aes(label = fmt_n(n)), hjust = -0.15, size = 2.8) +
  scale_x_continuous(labels = label_whole, expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Owner type 2020 → 2026",
    x = "Parcels",
    y = NULL
  ) +
  theme_one


```


<br>


```{r}
#| label: fig-delta-owner
#| fig-cap: "Ownership name continuity among major-loss parcels."


name_df <- loss |>
  group_by(name_from, name_to) |>
  summarise(n = n(), acres = sum(acres_2020), .groups = "drop") |>
  mutate(flow = reorder(paste(name_from, "→", name_to), acres))

ggplot(name_df, aes(acres, flow)) +
  geom_col(fill = "#8b5a2b", width = 0.7) +
  geom_text(
    aes(label = paste0(fmt_n(n), " · ", fmt_n(acres))),
    hjust = -0.05,
    size = 2.6
  ) +
  scale_x_continuous(labels = label_whole, expand = expansion(mult = c(0, 0.28))) +
  labs(
    title = "Owner name 2020 → 2026",
    x = "Acres",
    y = NULL
  ) +
  theme_one
```

### Canopy cover change

```{r}
#| label: fig-delta-type
#| fig-cap: "Distribution of canopy change (percent) for major-loss parcels."
#| fig-pos: "H"
ggplot(loss, aes(delta_pp)) +
  geom_histogram(binwidth = 5, fill = "#1f4e3a", color = "white", linewidth = 0.3) +
  geom_vline(xintercept = -20, linetype = 2, color = "grey40") +
  scale_x_continuous(breaks = seq(-75, -20, by = 10), labels = label_whole) +
  scale_y_continuous(labels = label_whole) +
  labs(
    title = "How large were the losses?",
    subtitle = "Mean parcel TCC (percent); dashed line = −20 percent threshold",
    x = "Change in canopy (2025 − 2020), percent",
    y = "Parcels"
  ) +
  theme_one
```

```{=latex}
\FloatBarrier
```

### Takeaway

The ~`{r} fmt_n(ac_loss)` acres of **major canopy loss** are concentrated on **forest-industry CFP land that kept the forest industry type label**. Continuity of owner *type* masks substantial **legal-name turnover**, especially **TRG Threshold Timber → Verdant Timber** (~55 parcels / ~`{r} fmt_n(ac_trg_verdant)` ac) and **Keweenaw Land Association → Sage Timber** (8 parcels / ~`{r} fmt_n(ac_kla_sage)` ac). None of these `{r} fmt_n(n_loss)` parcels left the CFP between 2020 and 2026.

```{=latex}
\FloatBarrier
```

### Methods (short)

- **Unit:** 2020 CFP parcels with attribute acres ≥ 20 (`parid`).
- **Canopy:** zonal mean USFS/NLCD TCC on the 30 m grid, **2020 vs 2025** (latest annual TCC; ownership snapshot is 2026).
- **Major loss:** mean ΔTCC < −20 percent (`{r} fmt_n(n_loss)` of `{r} fmt_n(n_all)` ranked parcels).
- **Ownership:** 2020 `owner_type` / `search_nam`; 2026 `OwnerTypeT` / `FullLegalN`, joined on `parid`.
- **Name match:** loose string compare (case, punctuation, LLC/Inc stripped). Source table: `data/processed/cfp_parcel_tcc_change_by_owner_type.csv` (`scripts/09_cfp_parcel_tcc_change.R`).
